Prediction of the 2020 US Presidential Election using MRP on a non-representative poll

### 2020-11-03 outputs/paper/paper.Rmd
final commit

---
title: "Prediction of the 2020 Presidential Election"
subtitle: "Spoiler Alert: Trump Wins"
author: "Anees Shaikh, Jaffa Romain, Lu Mu, and Cameron Fryer"
thanks: "Code and data are available at: https://github.com/aneesshake/elections_prediction"
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: |
  | The purpose of this analysis is to forecast the results of the upcoming United States 2020 presidential election. In this paper, we first consider the Democracy Fund + UCLA Nationscape Wave 50 dataset, which contains the results of a survey (conducted June 25-July 1, 2020) on American voter attitudes. Thereafter, the survey data is used to train our model relating voter intent to 4 explanatory variables. The model is then applied to the post-stratification dataset; namely, the results of the 2018 1-year American Community Survey (ACS), to get predictions representative of the target population. Since the ACS data pertains to individual persons and their characteristics, the associated use of our model allows us to conclude that Donald Trump will be relected as the President of the United States, winning the popular vote by ___ with a margin of error of ______
Keywords: forecasting; US 2020 election; Trump; Biden; multilevel regression with post-stratification
output:
  bookdown::pdf_document2:
    citation_package: natbib
toc: FALSE
bibliography: references.bib
---

@@

There are a vast number of ways in which statistical analysis can be sued to foretell future political future  political appointment. With that being said, however, the statistical methods one chooses to exploit are usually dependent on the data in question. As such, it is important to both understand, and be able to describe the data before considering the optimal approach in working with it. By staying true to this procedure deriving inferences about some given 
# Data

@@

As previously mentioned, our data is from two different datasets; the Democracy Fund + UCLA Nationscape Wave 50 dataset, and the 2018 1-year American Community Survey (ACS) dataset. They each contain both qualitative and quantitative data. 
To forecast the results of the presidential election, our model was fitted using the Nationscape survey data. The Democracy Fund + UCLA Nationscape is "one of the largest public opinion surveys ever conducted -- interviewing people in nearly every county, congressional district, and mid-sized U.S. city in the lead up to the 2020 election (Tausanovitch, 2020)." Consequently, interviewing "roughly 6,250 per week" was part of the survey methodology employed by Nationscape. As such, the Nationscape Wave 50 dataset contains results for 6,479 interviews conducted during the week of June 25-July 1, 2020. It can be accessed by visiting https://www.voterstudygroup.org/publication/nationscape-data-set and submitting a request to access the data (Tausanovitch, 2020). By providing your full name and email address, you will then be emailed a link to the page where it can be downloaded.

Nationscape samples are provided by Lucid, a market research platform that uses their software for conducting online surveys. Nationscape surveys were completed by participants online in English (Tausanovitch, 2020). The survey data is then weighted using a simple raking technique, where one set of weights is generated for each week's survey, in an attempt to have data that is representative of their target population. The survey data is weighted on many demographic factors such as gender, age, and nativity.

  At first glance of the Wave 50 data, we noted that it contains 265 variables. This is noteworthy because it allows a more complete picture of voter attitude to be derived from the data compared to if there were less variables. Another strength of this dataset is the overall quality of responses from those surveyed. More specifically, the data does not contain any duplicate records, inappropriate entries, or poor entries (e.g. typos, misspellings); hence, making it easier to work with. This high quality of survey responses can be at least partially attributed to the fact that "Nationscape samples are provided by Lucid, a market research platform that runs an online exchange for survey respondents(Tausanovitch, 2020)". What is more, the individuals belonging to the sample are required to "complete an attention check before completing the [online] survey (Tausanovitch, 2020)." With this method of sample selection, along with the precaution taken to ensure their attentiveness, it is unlikely for survey respondents to submit low-quality responses.

  The main weakness of the Nationscape Wave 50 dataset is the likelihood that it includes sampling error and/or coverage error. While the goal of the Nationscape survey is to determine American voter attitudes, making inferences based only on the Wave 50 dataset would result in sample bias. This is because the target population is all American citizens age 18 years and older, whereas the sample frame is Lucid's online exchange for survey respondents, and the sample is a set of 6,479 individuals from said online exchange which match "a set of demographic criteria (Tausanovitch, 2020)." It is unlikely that a survey sample of 6,479 American citizens during a single week is representative of American voter attitudes as a whole. This may lead one to wonder why Nationscape chose to use the described sampling approach. Apart from the fact that the survey was conducted every week, the reason is trade-offs. For example, although there is potential coverage error in the data--a common result of online surveys--some trade-offs are a lesser rate of non-response, lower cost, convenience and accuracy. Whilst on the topic of non-response, it is worth mentioning that the variable with greatest amount of non-response in the Wave 50 data set is household income at roughly 5.5%. Nonetheless, this is a much smaller percentage of non-response in household income than is typically seen in political surveys, and Nationscape handles it by not weighting income for non-respondents. Similarly, a statement from Nationscape in its "Representativeness Assessment" implies an overall lack of concern for non-response in the data, as "previous evaluations of the samples Lucid provides have found them to be of high quality (Tausanovitch, 2020)."

@@

```{r model2020prediction}
post_strat_estimates %>% group_by(middle_ground_election_status) %>% tally() %>% rename(state_result = middle_ground_election_status, number = n) %>%  knitr::kable(caption = "Model Prediction for 2020")
```

@@

While American citizens play a significant role in electing the next President, it is safe to say that the results of the US federal election are watched closely by the entire world. It goes without saying that the results of this election will have implications around the world. To contextualize the results of our model and analysis, we need a quick and simplified lesson on the American electoral process. Unlike most democracies, such as Canada, the head of state is typically elected by a popular vote. In the US, however, the head of state is not elected by a popular vote. Instead, citizens vote for a representative, or an elector, that then elects the president. This group of electors forms the electoral college. The electoral college is tasked with electing the next President. If a candidate wins a state by even a small percent, they effectively win all the electors in the state.

with that background knowledge in mind, what do our poll results indicate? From Figure(\@ref(fig:survey_election_results)), we gather that Trump has a very slight lead. Multiple other polls have shown Biden to be in the lead(FiveThirtyEight, 2020). Now there are a few reasons as to why we may not want to use poll results to predict elections. Polls have of late garnered a lot of criticism due to their incorrect predictions of the 2016 Federal election results[@citeAapor]. Note that while they got the ultimate results wrong, they did get the popular vote prediction correct[@citeAapor].Nearly every national poll that year had Hillary Clinton winning, with some estimates of her having a 3% point lead[@citeAapor]. Ultimately, she won the popular vote by 2.1%. However, as mentioned previously, the US does not use a popular vote system. At the state level, the polls became less effective, indicating a Clinton lead again[@citeAapor]. State level polls do translate over to the electoral college a bit more effectively than national polls do. Another reason we may not want to overly rely on polls is yet another cautious tale from the 2016 election season. In that season, polls had over-represented college-graduates, a group that was more likely to vote for Clinton, thus tipping the scales even more in favour of Clinton[@citeAapor]. Capturing this over-representation is key to effectively predicting an election. As mentioned in the results section, some results were intentionally shown as their distributions and then their voting preferences. This was done to elucidate the effect of a difference in representation, akin to  the difference of representation of educated voters between polled data and actual populations. 
From figure(\@ref(fig:age)), there is a clear shift in the distribution of ages. The increased representation of ages 65+ would likely predict a Trump win. Our model coefficient results also indicate that someone being in the oldest age group is more likely to vote for Trump, though significance is only slightly off. With an effectively older population compared to the Nationscape polling data, we can conclude that we would be likely see more votes for Trump coming out of our ACS group. Similarly, when looking at sex, we observe that there are more women in our ACS dataset than there are men, which is the opposite of the distribution found in the Nationscape data. Our model coefficient results again suggest that men are more likely to vote for Trump, and since we have more women in the ACS data, we would expect a higher percentage of Biden votes. This is further supported by other research that suggests that Trump has lost favour with white suburban women, a group he previously was well perceived by[@conroy2020]. A lot of models during the 2016 election season suggested a Clinton win, one of the few models that predicted a Trump win was a model that used MRP[@2016election]. It is worth mentioning that MRP isn't some sort of magic bullet as a few other MRP models during that time still predicted a Clinton win within the electoral college[@2016election]. How do current models fair? The primary model suggests that Trump has a 91% change of winning, in spite of Trump trailing Biden in the polls[@norpoth_2020].

Overall, our model predicts a Trump majority. The model predicts a win in 32 states and a close call in 10 other states. However, it is important to again caution that this is still the popular vote.

@@

One key weakness in our model is the way we present the results with regards to

@@

>>>>

---
title: "Prediction of the 2020 US Presidential Election using MRP on a non-representative poll"
subtitle: "Predicted Trump win by 52% within a 5% margin of error"
author: "Anees Shaikh, Jaffa Romain, Lu Mu, and Cameron Fryer"
thanks: "Code and data are available at: https://github.com/aneesshake/elections_prediction"
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: |
  | The purpose of this analysis is to forecast the results of the upcoming United States 2020 presidential election. In this paper, we first consider the Democracy Fund + UCLA Nationscape Wave 50 dataset, which contains the results of a survey (conducted June 25-July 1, 2020) on American voter attitudes. Thereafter, the survey data is used to train our model relating voter intent to 4 explanatory variables. The model is then applied to the post-stratification dataset; namely, the results of the 2018 1-year American Community Survey (ACS), to get predictions representative of the target population. Since the ACS data pertains to individual persons and their characteristics, the associated use of our model allows us to conclude that Donald Trump will be relected as the President of the United States, winning 52% of the popular vote with a margin of error of +-5%.
  | **Keywords:** forecasting; US 2020 election; Trump; Biden; multilevel regression with post-stratification
Keywords: "[forecasting; US 2020 election; Trump; Biden; multilevel regression with post-stratification]"
output:
  bookdown::pdf_document2:
    citation_package: natbib
toc: FALSE
bibliography: references.bib
---

@@

There are a vast number of ways in which statistical analysis can be used to foretell future political appointment. With that being said, however, the statistical methods one chooses to exploit are usually dependent on the data in question. As such, it is important to both understand, and be able to describe the data before considering the optimal approach in working with it. By staying true to this procedure, deriving inferences about some given dataset(s) will be accomplished much more smoothly.

Since the overall goal of this paper is to predict the winner of the 2020 United States Presidential Election using the Democracy Fund + UCLA Nationscape Wave 50 dataset (Tausanovitch and Vavreck, 2020) and the 2018 1-year American Community Survey (ACS) dataset(Ruggles et al, 2020), the first step that we took was becoming familiar with both sets of data. Consequently, it became clear that the most effective way to achieve our goal involved employing multi-level regression with post-stratification. What brought us to this conclusion is that, although the Nationscape survey contains bias, the survey answers it contained seemed superior to those included in the ACS dataset.  Furthermore, the sample contained in the ACS data better represented our target population—Americans 18 years and older. Thus, we used the Nationscape data to train a model which would then be paired with weightings extracted from the ACS data.

The remainder of the paper is structured as follows: Section 2 provides a summary of the datasets that we used, and highlights their key features; Section 3 involves an explanation of the model we constructed; Section 4  reveals the outcome of our modelling process; and lastly,  Section 5 provides a discussion on our results and notable findings. The Github repo is linked in the appendix.

# Data

@@

As previously mentioned, our data is from two different datasets; the Democracy Fund + UCLA Nationscape Wave 50 dataset, and the 2018 1-year American Community Survey (ACS) dataset. They each contain both qualitative and quantitative data. 
To forecast the results of the presidential election, our model was fitted using the Nationscape survey data. The Democracy Fund + UCLA Nationscape is "one of the largest public opinion surveys ever conducted -- interviewing people in nearly every county, congressional district, and mid-sized U.S. city in the lead up to the 2020 election (Tausanovitch and Vavreck, 2020)." Consequently, interviewing "roughly 6,250 per week" was part of the survey methodology employed by Nationscape. As such, the Nationscape Wave 50 dataset contains results for 6,479 interviews conducted during the week of June 25-July 1, 2020. It can be accessed by visiting https://www.voterstudygroup.org/publication/nationscape-data-set and submitting a request to access the data (Tausanovitch and Vavreck, 2020). By providing your full name and email address, you will then be emailed a link to the page where it can be downloaded.

Nationscape samples are provided by Lucid, a market research platform that uses their software for conducting online surveys. Nationscape surveys were completed by participants online in English (Tausanovitch and Vavreck, 2020). The survey data is then weighted using a simple raking technique, where one set of weights is generated for each week's survey, in an attempt to have data that is representative of their target population. The survey data is weighted on many demographic factors such as gender, age, and nativity.

  At first glance of the Wave 50 data, we noted that it contains 265 variables. This is noteworthy because it allows a more complete picture of voter attitude to be derived from the data compared to if there were less variables. Another strength of this dataset is the overall quality of responses from those surveyed. More specifically, the data does not contain any duplicate records, inappropriate entries, or poor entries (e.g. typos, misspellings); hence, making it easier to work with. This high quality of survey responses can be at least partially attributed to the fact that "Nationscape samples are provided by Lucid, a market research platform that runs an online exchange for survey respondents(Tausanovitch and Vavreck, 2020)". What is more, the individuals belonging to the sample are required to "complete an attention check before completing the [online] survey (Tausanovitch and Vavreck, 2020)." With this method of sample selection, along with the precaution taken to ensure their attentiveness, it is unlikely for survey respondents to submit low-quality responses.

  The main weakness of the Nationscape Wave 50 dataset is the likelihood that it includes sampling error and/or coverage error. While the goal of the Nationscape survey is to determine American voter attitudes, making inferences based only on the Wave 50 dataset would result in sample bias. This is because the target population is all American citizens age 18 years and older, whereas the sample frame is Lucid's online exchange for survey respondents, and the sample is a set of 6,479 individuals from said online exchange which match "a set of demographic criteria (Tausanovitch and Vavreck, 2020)." It is unlikely that a survey sample of 6,479 American citizens during a single week is representative of American voter attitudes as a whole. This may lead one to wonder why Nationscape chose to use the described sampling approach. Apart from the fact that the survey was conducted every week, the reason is trade-offs. For example, although there is potential coverage error in the data--a common result of online surveys--some trade-offs are a lesser rate of non-response, lower cost, convenience and accuracy. Whilst on the topic of non-response, it is worth mentioning that the variable with greatest amount of non-response in the Wave 50 data set is household income at roughly 5.5%. Nonetheless, this is a much smaller percentage of non-response in household income than is typically seen in political surveys, and Nationscape handles it by not weighting income for non-respondents. Similarly, a statement from Nationscape in its "Representativeness Assessment" implies an overall lack of concern for non-response in the data, as "previous evaluations of the samples Lucid provides have found them to be of high quality (Tausanovitch and Vavreck, 2020)."

@@

```{r model2020prediction,echo=FALSE}
post_strat_estimates %>% group_by(middle_ground_election_status) %>% tally() %>% rename(state_result = middle_ground_election_status, number = n) %>%  knitr::kable(caption = "Model Prediction for 2020")
```

@@

While American citizens play a significant role in electing the next President, it is safe to say that the results of the US federal election are watched closely by the entire world. It goes without saying that the results of this election will have implications around the world. To contextualize the results of our model and analysis, we need a quick and simplified lesson on the American electoral process. Unlike most democracies, such as Canada, the head of state is typically elected by a popular vote. In the US, however, the head of state is not elected by a popular vote. Instead, citizens vote for a representative, or an elector, that then elects the president. This group of electors forms the electoral college. The electoral college is tasked with electing the next President. If a candidate wins a state by even a small percent, they effectively win all the electors in the state. With that background knowledge in mind, what do our poll results indicate? From Figure(\@ref(fig:survey_election_results)), we gather that Trump has a very slight lead. Multiple other polls have shown Biden to be in the lead(FiveThirtyEight, 2020). Now there are a few reasons as to why we may not want to use poll results to predict elections. Polls have of late garnered a lot of criticism due to their incorrect predictions of the 2016 Federal election results[@citeAapor]. Note that while they got the ultimate results wrong, they did get the popular vote prediction correct[@citeAapor].Nearly every national poll that year had Hillary Clinton winning, with some estimates of her having a 3% point lead[@citeAapor]. Ultimately, she won the popular vote by 2.1%. However, as mentioned previously, the US does not use a popular vote system. At the state level, the polls became less effective, indicating a Clinton lead again[@citeAapor]. State level polls do translate over to the electoral college a bit more effectively than national polls do. Another reason we may not want to overly rely on polls is yet another cautious tale from the 2016 election season. In that season, polls had over-represented college-graduates, a group that was more likely to vote for Clinton, thus tipping the scales even more in favour of Clinton[@citeAapor]. Capturing this over-representation is key to effectively predicting an election. As mentioned in the results section, some results were intentionally shown as their distributions and then their voting preferences. This was done to elucidate the effect of a difference in representation, akin to  the difference of representation of educated voters between polled data and actual populations. From figure(\@ref(fig:age)), there is a clear shift in the distribution of ages. The increased representation of ages 65+ would likely predict a Trump win. Our model coefficient results also indicate that someone being in the oldest age group is more likely to vote for Trump, though significance is only slightly off. With an effectively older population compared to the Nationscape polling data, we can conclude that we would be likely see more votes for Trump coming out of our ACS group. Similarly, when looking at sex, we observe that there are more women in our ACS dataset than there are men, which is the opposite of the distribution found in the Nationscape data. Our model coefficient results again suggest that men are more likely to vote for Trump, and since we have more women in the ACS data, we would expect a higher percentage of Biden votes. This is further supported by other research that suggests that Trump has lost favour with white suburban women, a group he previously was well perceived by[@conroy2020]. A lot of models during the 2016 election season suggested a Clinton win, one of the few models that predicted a Trump win was a model that used MRP[@2016election]. It is worth mentioning that MRP isn't some sort of magic bullet as a few other MRP models during that time still predicted a Clinton win within the electoral college[@2016election]. How do current models fair? The primary model suggests that Trump has a 91% change of winning, in spite of Trump trailing Biden in the polls[@norpoth_2020]. Overall, our model predicts a Trump majority. The model predicts a win in 32 states and a close call in 10 other states. However, it is important to again caution that this is still the popular vote.

@@

One key weakness in our model is the way we present the results with regards to states won. If we look at it from a conservative estimate point of view, where we only use the lower bounds, then we'll see a vastly different election result. We chose to remove state from the model as nearly no state had a significant value in its coeffiecents. This model could be enhanced by using a different poll. For instance @2016_election uses an A+ quality survey, as determined by FiveThirtyEight, and was able to best predict the 2016 election. We could try that approach and see if the result changes. The number of dimensions we have are too few, there were a few variables within the Nationscape dataset such as voting behaviour in 2016 that we weren't able to get a similar value for from the post-stratified dataset. Such variables likely have high explanatory power and would make our model overall much more effective. A 5% margin of error is quite large, especially when elections are extremely close. A reduction in the margin of error would result in being able to much more accurately predict the election.

@@

All code can be found at: https://github.com/aneesshake/elections_prediction
